### 2021-07-09 newvariant.tex
edits

Inferring the relative strength (i.e., the ratio of reproduction numbers, $\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
Although the role of the generation interval (i.e., time between infection and transmission) in linking epidemic strength and speed has been previously well-established, many analyses of new SARS-CoV-2 variants overlook uncertainties in the generation-interval distributions as well as the possibility that their inferences of epidemic dynamics can depend on whether relative strength or speed are held constant.
Here, we explore how differences in the generation-interval distributions affect the inference of relative strength and speed in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
We use simulations to show how neglecting these differences can lead to biases in estimates of relative strength in practice and to illustrate how such biases can be assessed.
We also assess which of the two assumptions (i.e., constant relative strength or speed) are more appropriate and further discuss implications for control.

@@

Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
Instead, epidemic strength can be inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
The generation interval is defined as the time between infection and transmission and provides information about the individual-level transmission \citep{svensson2007note}.
The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength is fixed, assuming longer generation intervals lead to a slower epidemic (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on hazard of isolation and therefore affects late transmission disproportionately.
When epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contact and therefore affecting transmission by an equal amount regardless of age of infection.
Therefore, it is critical to evaluate which assumption is more appropriate for inferring transmission advantages of new variants.

In addition, most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
However, linking strength and speed is complicated given that the generation-interval distribution depend on many factors including behavior: 
for example, self-isolation after symptom onset will lead to shorter generation intervals.

Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.

>>>>

Inferring the relative strength (i.e., the ratio of reproduction numbers, $\rho=\Rv/\Rw$) and relative speed (i.e., the difference between growth rates, $\delta=\rv-\rw$) of new SARS-CoV-2 variants compared to their wild types is critical to predicting and controlling the course of the current pandemic.
We explore how differences in the generation-interval distributions affect the inference of relative strength $\rho$ and speed $\delta$ in light of recent evidence that the B.1.1.7 variant may have a longer infectious period (and therefore, a longer generation interval) than prior dominant lineages.
Neglecting these differences gives biased estimates of $\rho$, which can vary systematically with changes in the incidence of infections caused by the variant.
We further argue that holding $\rho$ constant is generally more appropriate in the context of SARS-CoV-2 and illustrate how $\rho$ and $\delta$ are affected by two idealized interventions.
A constant-strength intervention, which reduces transmission rates by a constant amount, affects $\delta$ but not $\rho$---this is analogous to vaccination and social distancing.
A constant-speed intervention, which isolates infected individuals at a constant hazard, affects $\rho$ but not $\delta$---this is analogous to contact tracing.
Uncertainties in the generation-interval distribution determine the effectiveness of these interventions.

@@

Epidemic speed can be typically estimated from incidence of infection during the exponential growth period \citep{mills2004transmissibility,nishiura2009transmission,ma2014estimating}, but epidemic strength is difficult to measure directly.
Instead, epidemic strength can be inferred from the observed epidemic speed using the generation-interval distribution, an approach popularized by \citep{wallinga2007generation}.
The generation interval is defined as the time between infection and transmission and provides information about the time scale of individual-level transmission \citep{svensson2007note}.
The exact shape of the distribution depends on several factors---including the shape of latent and infectious periods \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model} as well as more detailed life history of a disease \citep{huber2016quantitative}---and thus can be difficult to estimate.
While it is possible to consider general forms of generation-interval distributions \citep{miller2010epidemics,svensson2015influence}, summarizing the distribution in terms of its mean and variability---for example, by using the Gamma distribution---can still provide a robust link between epidemic speed and strength for real diseases and yield important biological insights \citep{park2019practical}.
In particular, several studies have noted, in many contexts, that mechanisms that increase the mean generation interval can increase epidemic strength given observed epidemic speed \citep{eaton2014proportion,powers2014impact,weitz2015modeling,park2020time}.

Analyses of new variants have typically characterized \emph{relative} strength (i.e., the ratio of reproduction numbers $\Rv/\Rw$) and speed (i.e., the difference between growth rates $\rv-\rw$) of the variants.
Many studies have tried to estimate the relative strength of variants from the observed relative speed \citep{davies2021estimated, leung2021early, volz2021transmission,zhao2021,campbell2021increased}.
Some studies have instead assumed a value of the relative strength and tried to predict its relative speed \citep{davies2021estimated,di2021impact}.
While both approaches are reasonable, holding different quantities constant (i.e., strength or speed) can lead to different conclusions about the spread of the disease and its control because two quantities are linked by the generation-interval distribution \citep{doi:10.1098/rspb.2020.1556}.
For example, when epidemic strength $\RR$ is fixed, assuming longer generation intervals leads to a slower epidemic growth (lower $r$), making the epidemic look easier to control with speed-like interventions (e.g., contact tracing), which depend on isolation of infected individuals and therefore affect late transmission disproportionately---as shown in \cite{doi:10.1098/rspb.2020.1556}, the effective hazard of isolation must be greater than the epidemic speed $r$ to reach the $r=0$ threshold.
However, since $\RR$ is fixed, the generation-interval distribution is expected to have negligible effects on the effectiveness of strength-like interventions (e.g., social distancing), which reduce the hazard of infectious contacts and therefore affect transmission by an equal amount regardless of age of infection.
Conversely, when epidemic speed is fixed, assuming longer generation intervals lead to a stronger epidemic (higher $\RR$), making the epidemic look harder to control with strength-like interventions (but not necessarily with speed-like interventions).

Most studies have assumed that the previous dominant strain (wild type) and variants have identical generation-interval distributions, but recent evidence suggests that the B.1.1.7 variant may have a longer duration of infection: 13.3 days (90\% CI: 10.1--16.5) for the new variant and 8.2 days (90\% CI: 6.5--9.7) for the wild type \citep{kissler2021densely}.
Longer duration of infection suggests that the mean generation interval of B.1.1.7 is likely to be longer than that of the wild type \citep{lloyd2001realistic,wearing2005appropriate,roberts2007model}.
Other studies have considered the possibility that the faster growth rate of new variants may be driven, in part, by shorter generation intervals \citep{davies2021estimated,volz2021transmission}.
However, linking strength and speed is complicated given that the generation-interval distribution depend on many factors including behavior: 
for example, self-isolation after symptom onset will lead to shorter generation intervals.

Here, we explore how different assumptions affect estimates of the relative strength (the ratio $\rho=\Rv/\Rw$) and relative speed (the difference $\delta=\rv-\rw$) for a new variant, such as B.1.1.7.
In doing so, we assess 
% We compare $\rho$--$\delta$ relationship between relative strength and relative speed under a wide range of assumptions about generation-interval distributions.
We show that neglecting differences in the generation-interval distributions can lead to biased estimates.
We also discuss how such biases might be assessed in practice and how information on differences in generation interval distributions might influence priorities for controlling the spread of VoCs.
